---
title: "Analysis of Biden approval margin"
author: "Study Group: 12 (Akshat Kacheria, Asli Ceren Memis, Jelena Savic, Shengcheng Liu, Sophia Kalusche)"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(infer)
```





```

# Bidens Approval Margins

As we saw in class, fivethirtyeight.com has detailed data on [all polls that track the president's approval ](https://projects.fivethirtyeight.com/biden-approval-ratings)

```{r, cache=TRUE}

# Import approval polls data directly off fivethirtyeight website
approval_pollist <- read_csv('https://projects.fivethirtyeight.com/biden-approval-data/approval_polllist.csv') 

glimpse(approval_pollist)

# Use `lubridate` to fix dates, as they are given as characters.
approval_pollist <- approval_pollist %>% 
  mutate(modeldate = lubridate::mdy(modeldate), 
         startdate = lubridate::mdy(startdate), 
          enddate = lubridate::mdy(enddate), 
          createddate = lubridate::mdy(createddate))
```

## Create a plot

What I would like you to do is to calculate the average net approval rate (approve- disapprove) for each week since he got into office. I want you plot the net approval, along with its 95% confidence interval. There are various dates given for each poll, please use `enddate`, i.e., the date the poll ended.

```{r Bidens Approval Margins, fig.width=12, fig.height=6}

# Create confidence levels
approval_margins <- approval_pollist %>%
  
  #Select enddate
  filter(!is.na(enddate)) %>%
  mutate(week=isoweek(enddate),
         margin=approve-disapprove) %>%
  
  #Group the data
  group_by(week, subgroup) %>%
  
  #Summarize data (use se formula for differences)
  summarise(
    mean=mean(margin), 
    sd=sd(margin), 
    count=n(), 
    se=sd/sqrt(count), 
    t_critical=qt(0.975, count-1), 
    lower=mean-t_critical*se, 
    upper=mean+t_critical*se)

glimpse(approval_margins)


#Create the graph
approval_margins %>% 
  filter(subgroup == "Voters") %>%
  ggplot(aes(x=week, y=mean)) +
  
  #Set colors 
  geom_point(color="chocolate2", size=1.5) +
  geom_line(color="chocolate2")+
  
  #Add fill between lines
  geom_ribbon(aes(ymin=lower, ymax=upper),
              color="chocolate2", 
              fill="grey87", 
              linetype=1, 
              alpha=0.5, 
              size=0.3) +
  
  #Change limits, theme, scale, facet wrap and add fitted line 
  ylim(c(-15,50)) +
  theme_bw() +
  scale_x_continuous(breaks=seq(0, 35, 13))+
  scale_y_continuous(breaks=seq(-15, 10, 2.5))+
  geom_smooth(se=FALSE) +
  
  #Add horizontal line 
  geom_hline(yintercept=0, 
             linetype="solid",
             color = "chocolate2", 
             size=2) +
  
  #Add labels 
  labs( title="Estimating Approval Margins (approve-disapprove) for Joe Biden",
        subtitle = "Weekly average of all polls",
        x = "Week of the year",
        y = "Average Approval Margin (Approve - Disapprove)") +
  NULL
    

```



## Compare Confidence Intervals

Compare the confidence intervals for `week 4` and `week 25`. Can you explain what's going on? One paragraph would be enough.

The sample set differs between week 4 and week 25. The sample size on week 3 is much smaller than that of week 25 which is why the standard error is relatively higher in week 3. This leads to larger confidence intervals in week 3 compared to week 25.
As far as the data across the weeks is concerened, as sample size increases - confidence intervals shrink. The approval ratings for Joe Biden have reduced between week 4 and 25. 


